---
title: "Giovanna_data_2_all_rarefied"
author: "Nicholas Chance"
date: "2025-12-10"
output: 
  html_document:

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width = 15, fig.height=9)
```

# Library
```{r}
#open packages necessary#
library(phyloseq) #needed to create phyloseq object#
library(csv) #to read csv files, in this case GMeta from excel#
library(tidyverse) #cleans up large data sets/tables#
library(ggplot2) #plotting#
library(vegan) #microbial comm analysis, diversity indices#
library(ggvegan) #plotting specifically with vegan#
library(ggsci) #more/larger color palettes# 
library(randomcoloR) #specifically for distinct color palettes)
library(igraph) #graphing network analysis#
library(MiscMetabar)
library(ggVennDiagram) #make venn diagrams#
library(cowplot)
library(RColorBrewer)
library(ggpubr)
library(patchwork)
library(RRPP)
library(maaslin3)
library(multcompView)
library(pairwiseAdonis)
```

## open files
```{r}
#Open files#
taxonomy_giovanna <- 
  readRDS("C:/Users/nicho/OneDrive/SeqCenter/Giovanna_16S Analysis/Giovanna_all_days/Giovanna1.taxa.rds")
seqtab_nochim_giovanna <-
  readRDS("C:/Users/nicho/OneDrive/SeqCenter/Giovanna_16S Analysis/Giovanna_all_days/Giovanna1.seqtab.nochim.rds")

#Convert files into useable data frames#
Taxtab <- as.data.frame(taxonomy_giovanna)
ASVtab <- as.data.frame(seqtab_nochim_giovanna)
#open metadata#
GMeta <- as.csv("C:/Users/nicho/OneDrive/SeqCenter/Giovanna_16S Analysis/Giovanna_all_days/G_metadata_all_2.csv", 
                row.names = 1, header = TRUE, sep = ",", 
                check.names = TRUE, stringsAsFactors = TRUE)
```

## Standardize variables in metadata
```{r}
GMeta_std<-GMeta
numeric_col<-c("OLR_num", "pH","Ammonia", "TVFA")
GMeta_std[numeric_col]<-vegan::decostand(GMeta_std[numeric_col], "standardize")
```

### check if it worked
```{r}
# check column means (to 1 decimal place)- should be zero
round(apply(GMeta_std[numeric_col], 2, mean), 1)
```
```{r}
# check std dev of columns- should be 1
apply(GMeta_std[numeric_col], 2, sd)
```
## base phyloseq object
```{r}
samdata_base<-sample_data(GMeta)
GV_ps_base = phyloseq(otu_table(seqtab), tax_table(taxtab), sample_data(samdata_base))
```


## Rarefy object
```{r}
#rarefy ASVtab with vegan rarefaction# 
ASVtab_rar_veg <- rrarefy(ASVtab, sample = 4726)

ASVtab[1:3,1:3]
colnames(ASVtab) <- NULL

#naming for phyloseq object#
samdata <- sample_data(GMeta_std)
seqtab <- otu_table(ASVtab, taxa_are_rows = FALSE)
taxtab <- tax_table(Taxtab)
seqtab_veg <- otu_table(ASVtab_rar_veg, taxa_are_rows = FALSE)
#Check that names match to output#
head(sample_names(samdata))
head(sample_names(seqtab))
head(taxa_names(seqtab))
head(taxa_names(taxtab))

#### rarefy asv table####
seqtab_rar <- rarefy_even_depth(seqtab, sample.size = min(sample_sums(seqtab)), 
                                rngseed=5, trimOTUs=TRUE, replace = FALSE )

#making phyloseq object#
GV_all_ps = phyloseq(otu_table(seqtab), tax_table(taxtab), sample_data(samdata))

#make phyloseq object with otu table rarefied first
GV_all_rar_1 = phyloseq(otu_table(seqtab_rar), tax_table(taxtab), 
                        sample_data(samdata))
sample_sums(GV_all_rar_1)
####rarefy whole phyloseq object####
GV_all_rar <- rarefy_even_depth(GV_all_ps, 
                                sample.size = min(sample_sums(GV_all_ps)), 
                                                             rngseed=5, 
                                trimOTUs=TRUE, replace = FALSE)
GV_all_rar_d <-psmelt(GV_all_rar)
GV_genus_rar <-tax_glom(GV_all_rar, taxrank = "ta6")
GV_genus_rar_d <- psmelt(GV_genus_rar)


####rarefy with vegan rarefy####
GV_all_rar_veg <-phyloseq(otu_table(seqtab_veg), tax_table(taxtab), 
                          sample_data(samdata))
sample_sums(GV_all_rar_veg)

GV_all_rar_veg_d <-psmelt(GV_all_rar_veg)
GV_genus_rar_veg <-tax_glom(GV_all_rar_veg, taxrank = "ta6")
GV_genus_rar_veg_d <- psmelt(GV_genus_rar_veg)
```

## rarefaction curve
```{r}
S <- specnumber(ASVtab_rar_veg)
(raremax<-min(rowSums(ASVtab_rar_veg)))
Srare <- rarefy(ASVtab_rar_veg, raremax)
rarecurve(ASVtab_rar_veg, step = 50, sample = raremax, col = "blue", cex = .2)
```


# Relative abundance
```{r}
GV_ps_percentage_rar = phyloseq::transform_sample_counts(GV_genus_rar, 
                                                     function(x){x / sum(x)})
GV_ps_relative_rar_d <- psmelt(GV_ps_percentage_rar)
```

```{r}
GV_ps_relative_rar_d$ta6 <- as.character(GV_ps_relative_rar_d$ta6)
GV_ps_relative_rar_d1 <- GV_ps_relative_rar_d %>%
  group_by(Day_ID, substrate, ta6) %>%
  mutate(median=median(Abundance))
keep <- unique(GV_ps_relative_rar_d1$ta6[GV_ps_relative_rar_d1$median >.01])
GV_ps_relative_rar_d1$ta6[!(GV_ps_relative_rar_d1$ta6 %in% keep)] <- "< 1%"
GV_ps_rel_sum_rar <- GV_ps_relative_rar_d1%>%
  group_by(Sample,Day_ID, substrate, ta6) %>%
  summarise(Abundance=sum(Abundance))
```

```{r}
load("C:/Users/nicho/OneDrive/Documents/GV_palette_ra.RData") # previously saved palette specifically for this plots
```

# all substrate relative abundance graph
```{r}
GV_ps_rel_sum_rar$Day_ID <-
  factor(GV_ps_rel_sum_rar$Day_ID,
         levels=c("0-1", "0-2", "16-1","16-2","24-1","24-2","32-1","32-2",
                  "36-1","36-2","40-1","40-2","44-1","44-2","52-1","52-2",
                  "60-1","60-2","72-1","72-2",
                  "88-1","88-2","100-1","100-2","120-1","120-2","128-1","128-2",
                  "132-1","132-2","140-1","140-2","156-1","156-2"))
GV_ps_rel_sum_rar$substrate <-
  factor(GV_ps_rel_sum_rar$substrate, 
         levels=c("Protein","Carbohydrate","Grease"))

#Plot final relative abundance chart#
ggplot(GV_ps_rel_sum_rar, aes(x = Day_ID, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=ta6)) +
  labs(x="", y="% Abundance") +
  facet_wrap(~substrate, scales= "free_x", nrow=1) +
  scale_fill_manual(values = (GV_palette_ra))+
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.title.y = element_text(size=12),
        axis.text.x = element_text(size=15),
        strip.text.x = element_text(size=20),
        legend.key.size = unit(.7,'cm'),
        axis.text.x.bottom = element_text(angle = 90),
        legend.position = "bottom",
        legend.direction = "horizontal",
        legend.text=element_text(size=12),
        plot.margin = unit(c(0,0,0,0),'cm'),
        legend.margin = margin(l=-0.05, unit="npc"))+
  guides(fill = guide_legend(nrow = 9))
```

# Subset samples by domain 
```{r}
GV_archaea_rar <- subset_taxa(GV_genus_rar, ta1 == "Archaea" )
GV_arcahaea_rar_d <- psmelt(GV_archaea_rar)

GV_bacteria_rar <- subset_taxa(GV_genus_rar, ta1 == "Bacteria" )
GV_bacteria_rar_d <- psmelt(GV_bacteria_rar)
```

# Subset samples by substrate
```{r}
#base
GV_base_p<-subset_samples(GV_ps_base, substrate == "Protein")
GV_base_c<-subset_samples(GV_ps_base, substrate == "Carbohydrate")
GV_base_g<-subset_samples(GV_ps_base, substrate == "Grease")
#standardized
GV_ps_protein_rar <-subset_samples(GV_genus_rar, substrate == "Protein")
GV_ps_carb_rar <-subset_samples(GV_genus_rar, substrate == "Carbohydrate")
GV_ps_grease_rar <- subset_samples(GV_genus_rar, substrate == "Grease")
```

## ASVtab and samdata from each substrate subset (as data frame)
```{r}
#base protein
samdata_p_std<-data.frame(GV_base_p@sam_data)
samdata_p_std[numeric_col]<-vegan::decostand(samdata_p_std[numeric_col], "standardize")

#base carb
samdata_c_std<-data.frame(GV_base_c@sam_data)
samdata_c_std[numeric_col]<-vegan::decostand(samdata_c_std[numeric_col], "standardize")

#base carb
samdata_g_std<-data.frame(GV_base_g@sam_data)
samdata_g_std[numeric_col]<-vegan::decostand(samdata_g_std[numeric_col], "standardize")

#protein
ASVtab_p<-as.data.frame(GV_ps_protein_rar@otu_table)
samdata_p<-as.data.frame(GV_ps_protein_rar@sam_data)
#carbohydrate
ASVtab_c<-as.data.frame(GV_ps_carb_rar@otu_table)
samdata_c<-as.data.frame(GV_ps_carb_rar@sam_data)
#grease
ASVtab_g<-as.data.frame(GV_ps_grease_rar@otu_table)
samdata_g<-as.data.frame(GV_ps_grease_rar@sam_data)



```


# PCoA on all Bacteria
```{r}
bray_pcoa_rar_b_1 <- phyloseq::ordinate(
  physeq = GV_bacteria_rar,
  method = "PCoA", 
  distance = "bray")

sample_data(GV_bacteria_rar)$OLR_VS <- factor(sample_data(GV_bacteria_rar)$OLR_VS,
         levels=c("Start","Inc_1","Inc_2","Inc_3","Inc_4","Recovery"))
P_all_b_pcoa<- plot_ordination(
  physeq = GV_bacteria_rar,          
  ordination = bray_pcoa_rar_b_1, label = "Day_ID")+ 
  geom_text(label = "", check_overlap=TRUE)+ 
  geom_point(aes(shape = substrate, 
                 color = as.factor(OLR_VS)), size = 3, stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","black","orange"))+
  scale_shape_manual(values = c(15,16,17))+
  guides(fill = guide_legend(override.aes = list(shape = 21)))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  stat_ellipse(type="t", aes(group=substrate,linetype=substrate))+
  labs(title = "D")
```


# Canonical Correspondance analysis (CCA)
```{r}
#### Canonical Correspondence analysis- protein####
cca_protein_rar <- phyloseq::ordinate(GV_ps_protein_rar, "CCA", 
                        formula = GV_ps_protein_rar ~OLR_num+pH+Ammonia+TVFA)
print(cca_protein_rar) # allow for checking inertia/fit#
cca_p_p <- plot(cca_protein_rar, display = c("wa","bp","cn"))

####Canonical Correspondence analysis- carbohydrate####
cca_carb_rar <- phyloseq::ordinate(GV_ps_carb_rar, "CCA", 
                     formula = GV_ps_carb_rar ~OLR_num+pH+Ammonia)
print(cca_carb_rar) # allow for checking inertia/fit#
plot(cca_carb_rar, display =c("wa","bp","cn"))

####Canonical Correspondence analysis- grease####
cca_grease_rar <- phyloseq::ordinate(GV_ps_grease_rar, "CCA",
                       formula = GV_ps_grease_rar ~OLR_num+pH+Ammonia+TVFA)
print(cca_grease_rar)
cca_g_p<-plot(cca_grease_rar, display =c("wa","bp","cn"))

(cca_p_p | cca_c_p)
```

# Alpha Diversity
```{r}
#### alpha diversity####
d_richness_rar <- estimateR(seqtab_rar) #richness and chao1#
#d_evenness_rar <- diversity(seqtab_rar) / log(specnumber(ASVtab_rar_veg))  #evenness#
d_shannon_rar <- diversity(seqtab_rar, index= "shannon") #shannon#
d_simpson_rar <- diversity(seqtab_rar, index = "simpson")
d_alphadiv_rar <- cbind(GMeta, t(d_richness_rar), d_shannon_rar, d_simpson_rar) #combine into 1 table#

rm(d_richness, d_shannon, d_simpson) 
d_alphadiv_tidy_rar <- 
  d_alphadiv_rar %>%
  mutate(sample_id = rownames(d_alphadiv_rar)) %>%
  gather(key   = alphadiv_index,
         value = obs_values,
         -sample_id, -substrate, -OLR_VS)

#plot- for total substrate samples#
P_richness_rar <- ggplot(d_alphadiv, aes(x=substrate, y=S.obs)) +
  geom_boxplot(fill=c("blue","red","green")) +
  labs(title= 'Richness', x= ' ', y= '', tag = "A") +
  geom_point()

P_chao_rar <- ggplot(d_alphadiv, aes(x=substrate, y=S.chao1)) +
  geom_boxplot(fill=c("blue","red", "green")) +
  labs(title= 'Chao1', x= ' ', y= '', tag = "B") +
  geom_point()

P_simpson_rar <- ggplot(d_alphadiv, aes(x=substrate, y=d_simpson)) +
  geom_boxplot(fill=c("blue","red", "green")) +
  labs(title= 'Simpson', x= ' ', y= '', tag = "C") +
  geom_point()

P_shannon_rar <- ggplot(d_alphadiv, aes(x=substrate, y=d_shannon)) +
  geom_boxplot(fill=c("blue","red","green")) +
  labs(title= 'Shannon', x= ' ', y= '', tag = "D") +
  geom_point()

#combine#
(P_richness_rar | P_chao_rar) / (P_simpson_rar | P_shannon_rar)
```

# Protein
## Subset
```{r}
GV_archaea_p_rar <- subset_samples(GV_archaea_rar, substrate == "Protein")
GV_bacteria_p_rar <- subset_samples(GV_bacteria_rar, substrate == "Protein")
```

## Archaea relative abundance
```{r}
GV_archaea_p_rar_ra <- transform_sample_counts(GV_archaea_p_rar, 
                                           function(x) 100 * x/sum(x)) 
GV_archaea_p_rar_ra_d <- psmelt(GV_archaea_p_rar_ra)

GV_archaea_p_rar_ra_d$ta6 <- as.character(GV_archaea_p_rar_ra_d$ta6)
GV_archaea_p_rar_ra_d1 <- GV_archaea_p_rar_ra_d %>%
  group_by(Day_ID, OLR_VS, ta6) %>%
  mutate(median=median(Abundance))
keep <- unique(GV_archaea_p_rar_ra_d1$ta6[GV_archaea_p_rar_ra_d1$median >1])
GV_archaea_p_rar_ra_d1$ta6[!(GV_archaea_p_rar_ra_d1$ta6 %in% keep)] <- "< 1%"
GV_archaea_p_sum_rar <- GV_archaea_p_rar_ra_d1%>%
  group_by(Sample, Day_ID, OLR_VS, ta6) %>%
  summarise(Abundance=sum(Abundance)) 

GV_archaea_p_sum_rar$Day_ID <-
  factor(GV_archaea_p_sum_rar$Day_ID,
         levels=c("0-1", "0-2", "16-1","16-2","40-1","40-2","72-1","72-2",
                  "88-1","88-2","100-1","100-2","120-1","120-2",
                  "140-1","140-2","156-1","156-2"))
p_archaea_protein_rar <- distinctColorPalette(11)
ggplot(GV_archaea_p_sum_rar, aes(x = Day_ID, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=ta6)) +
  labs(x="", y="% Abundance") +
  scale_fill_manual(values = (p_archaea_protein_rar))+
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x.bottom = element_text(angle = 90))
```

## Bacteria relative abundance
```{r}
GV_bacteria_p_rar <- subset_samples(GV_bacteria_rar, substrate == "Protein")
GV_bacteria_p_rar_d <- psmelt(GV_bacteria_p_rar)
GV_bacteria_p_rar_ra <- transform_sample_counts(GV_bacteria_p_rar, 
                                            function(x) 100 * x/sum(x))
GV_bacteria_p_rar_ra_d <- psmelt(GV_bacteria_p_rar_ra)

GV_bacteria_p_rar_ra_d$ta6 <- as.character(GV_bacteria_p_rar_ra_d$ta6)
GV_bacteria_p_rar_ra_d1 <- GV_bacteria_p_rar_ra_d %>%
  group_by(Day_ID, OLR_VS, ta6) %>%
  mutate(median=median(Abundance))
keep <- unique(GV_bacteria_p_rar_ra_d1$ta6[GV_bacteria_p_rar_ra_d1$median >1])
GV_bacteria_p_rar_ra_d1$ta6[!(GV_bacteria_p_rar_ra_d1$ta6 %in% keep)] <- "< 1%"
GV_bacteria_p_sum_rar <- GV_bacteria_p_rar_ra_d1%>%
  group_by(Sample, Day_ID, ta6) %>%
  summarise(Abundance=sum(Abundance)) 


GV_bacteria_p_sum_rar$Day_ID <-
  factor(GV_bacteria_p_sum_rar$Day_ID,
         levels=c("0-1", "0-2", "16-1","16-2","40-1","40-2","72-1","72-2",
                  "88-1","88-2","100-1","100-2","120-1","120-2",
                  "140-1","140-2","156-1","156-2"))

p_bacteria_protein_rar <- distinctColorPalette(57)
ggplot(GV_bacteria_p_sum_rar, aes(x = Day_ID, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=ta6)) +
  labs(x="", y="% Abundance") +
  scale_fill_manual(values = (p_bacteria_protein_rar))+
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x.bottom = element_text(angle = 90))
```

# Carbohydrates
## Subset
```{r}
GV_archaea_c_rar <- subset_samples(GV_archaea_rar, substrate == "Carbohydrate")
GV_bacteria_c_rar <- subset_samples(GV_bacteria_rar, substrate == "Carbohydrate")
```

## Archaea relative abundance
```{r}
GV_archaea_c_ra_rar <- transform_sample_counts(GV_archaea_c_rar, 
                                           function(x) 100 * x/sum(x)) 
GV_archaea_c_ra_d_rar <- psmelt(GV_archaea_c_ra_rar)

GV_archaea_c_ra_d_rar$ta6 <- as.character(GV_archaea_c_ra_d_rar$ta6)
GV_archaea_c_ra_d1_rar <- GV_archaea_c_ra_d_rar %>%
  group_by(Day_ID, OLR_VS, ta6) %>%
  mutate(median=median(Abundance))
keep <- unique(GV_archaea_c_ra_d1_rar$ta6[GV_archaea_c_ra_d1_rar$median >1])
GV_archaea_c_ra_d1_rar$ta6[!(GV_archaea_c_ra_d1_rar$ta6 %in% keep)] <- "< 1%"
GV_archaea_c_sum_rar <- GV_archaea_c_ra_d1_rar%>%
  group_by(Sample, Day_ID, OLR_VS, ta6) %>%
  summarise(Abundance=sum(Abundance))

GV_archaea_c_sum_rar$Day_ID <-
  factor(GV_archaea_c_sum_rar$Day_ID,
         levels=c("0-1", "0-2", "16-1","16-2","24-1","24-2","36-1","36-2",
                  "44-1","44-2","60-1","60-2","88-1","88-2","100-1","100-2",
                  "128-1","128-2","140-1","140-2","156-1","156-2"))

p_archaea_carb_rar <- distinctColorPalette(12)
ggplot(GV_archaea_c_sum_rar, aes(x = Day_ID, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=ta6)) +
  labs(x="", y="% Abundance") +
  scale_fill_manual(values = (p_archaea_carb_rar))+
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x.bottom = element_text(angle = 90))
```

## Bacteria relative abundance
```{r}
GV_bacteria_c_d_rar <- psmelt(GV_bacteria_c_rar)
GV_bacteria_c_ra_rar <- transform_sample_counts(GV_bacteria_c_rar, 
                                            function(x) 100 * x/sum(x))
GV_bacteria_c_ra_d_rar <- psmelt(GV_bacteria_c_ra_rar)

GV_bacteria_c_ra_d_rar$ta6 <- as.character(GV_bacteria_c_ra_d_rar$ta6)
GV_bacteria_c_ra_d1_rar <- GV_bacteria_c_ra_d_rar %>%
  group_by(Day_ID, OLR_VS, ta6) %>%
  mutate(median=median(Abundance))
keep <- unique(GV_bacteria_c_ra_d1_rar$ta6[GV_bacteria_c_ra_d1_rar$median >1])
GV_bacteria_c_ra_d1_rar$ta6[!(GV_bacteria_c_ra_d1_rar$ta6 %in% keep)] <- "< 1%"
GV_bacteria_c_sum_rar <- GV_bacteria_c_ra_d1_rar%>%
  group_by(Sample, Day_ID, ta6) %>%
  summarise(Abundance=sum(Abundance))

GV_bacteria_c_sum_rar$Day_ID <-
  factor(GV_bacteria_c_sum_rar$Day_ID,
         levels=c("0-1", "0-2", "16-1","16-2","24-1","24-2","36-1","36-2",
                  "44-1","44-2","60-1","60-2","88-1","88-2","100-1","100-2",
                  "128-1","128-2","140-1","140-2","156-1","156-2"))

p_bacteria_carb_rar <- distinctColorPalette(58)
ggplot(GV_bacteria_c_sum_rar, aes(x = Day_ID, y = Abundance, fill = Genus)) + 
  geom_bar(stat = "identity", aes(fill=ta6)) +
  labs(x="", y="% Abundance") +
  scale_fill_manual(values = (p_bacteria_carb_rar))+
  theme_classic() +
  theme(strip.background = element_blank(),
        axis.text.x.bottom = element_text(angle = 90))
```

# Grease
## Subset
```{r}
GV_archaea_g_rar <- subset_samples(GV_archaea_rar, substrate == "Grease")
GV_bacteria_g_rar <- subset_samples(GV_bacteria_rar, substrate == "Grease")
```

#Beta Diversity- by substrate
##Protein
### PCoA Bacteria
```{r}
bray_pcoa_p_bacteria_rar <- phyloseq::ordinate(
  physeq = GV_bacteria_p_rar,
  method = "PCoA", 
  distance = "bray")

sample_data(GV_bacteria_p_rar)$OLR_VS <- 
  factor(sample_data(GV_bacteria_p_rar)$OLR_VS,
         levels=c("Start","Inc_1","Inc_2","Inc_3","Inc_4","Recovery"))
P_p_b_pcoa <- 
  plot_ordination(physeq = GV_bacteria_p_rar, ordination = bray_pcoa_p_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1, shape=17)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","black"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  stat_ellipse(type="t",linetype=1, aes(group=OLR_VS, color=OLR_VS))+
  labs(title = "A")
```

### PCoA Archara
```{r}
bray_pcoa_p_archaea_rar <- ordinate(
  physeq = GV_archaea_p_rar,
  method = "PCoA", 
  distance = "bray")

P_p_a_pcoa <- 
  plot_ordination(physeq = GV_archaea_p_rar, ordination = bray_pcoa_p_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Protein archaea PCoA rarefied")
```

### NMDS Archaea
```{r}
NMDS_p_archaea_rar <- ordinate(physeq = GV_archaea_p_rar, method = "NMDS", 
                           distance = "bray")

P_p_a_NMDS<- 
  plot_ordination(physeq = GV_archaea_p_rar, ordination = NMDS_p_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Protein archaea NMDS rarefied")
```

### NMDS Bacteria
```{r}
NMDS_p_bacteria_rar <- ordinate(physeq = GV_bacteria_p_rar, method = "NMDS", 
                               distance = "bray")

P_p_b_NMDS<-
  plot_ordination(physeq = GV_bacteria_p_rar, ordination = NMDS_p_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Protein NMDS bacteria rarefied")
```

### Combine plots
```{r}
(P_p_b_NMDS | P_p_b_pcoa 
  |PP_p_b_NMDS | PP_p_b_pcoa)/ (P_p_a_NMDS | P_p_a_pcoa | 
                                  PP_p_a_NMDS | PP_p_a_pcoa)
```

## Carbohydrate
### PCoA Bacteria
```{r}
bray_pcoa_c_bacteria_rar <- phyloseq::ordinate(
  physeq = GV_bacteria_c_rar,
  method = "PCoA", 
  distance = "bray")

sample_data(GV_bacteria_c_rar)$OLR_VS <- 
  factor(sample_data(GV_bacteria_c_rar)$OLR_VS,
         levels=c("Start","Inc_1","Inc_2","Inc_3","Recovery"))
P_c_b_pcoa<-
  plot_ordination(physeq = GV_bacteria_c_rar, ordination = bray_pcoa_c_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1, shape=15)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  stat_ellipse(type="t",linetype=1, aes(group=OLR_VS, color=OLR_VS))+
  labs(title = "B")
```

### NMDS Bacteria
```{r}
NMDS_c_bacteria_rar <- ordinate(physeq = GV_bacteria_c_rar, method = "NMDS", 
                            distance = "bray")

P_c_b_NMDS <- 
  plot_ordination(physeq = GV_bacteria_c_rar, ordination = NMDS_c_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Carbohydrate bacteria NMDS rarefied")
```

### PCoA Archaea
```{r}
pcoa_c_archaea_rar <- ordinate(
  physeq = GV_archaea_c_rar,
  method = "PCoA", 
  distance = "bray")

P_c_a_pcoa <-plot_ordination(physeq = GV_archaea_c_rar, ordination = pcoa_c_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Carbohydrate archaea PCoA rarefied")
```

### NMDS Archara
```{r}
NMDS_c_archaea_rar <- ordinate(physeq = GV_archaea_c_rar, method = "NMDS", 
                                distance = "bray")

P_c_a_NMDS <-plot_ordination(physeq = GV_archaea_c_rar, ordination = NMDS_c_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
labs(title = "Carbohydrate archaea NMDS rarefied")
```

### combine
```{r}
(P_c_b_NMDS | P_c_b_pcoa | 
    PP_c_b_NMDS | PP_c_b_pcoa) / (P_c_a_NMDS | P_c_a_pcoa |
                                    PP_c_a_NMDS | PP_c_a_pcoa)

```

## Grease
### PCoA Bacteria
```{r}
bray_pcoa_g_bacteria_rar <- phyloseq::ordinate(
  physeq = GV_bacteria_g_rar,
  method = "PCoA", 
  distance = "bray")

sample_data(GV_bacteria_g_rar)$OLR_VS <- 
  factor(sample_data(GV_bacteria_g_rar)$OLR_VS,
         levels=c("Start","Inc_1","Recovery","Inc_2","Inc_3"))
P_g_b_pcoa<-
  plot_ordination(physeq = GV_bacteria_g_rar, ordination = bray_pcoa_g_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","orange","red","pink"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  stat_ellipse(type="t",linetype=1, aes(group=OLR_VS, color=OLR_VS))+
  labs(title = "C")
```

### NMDS Bacteria
```{r}
NMDS_g_bacteria_rar <- ordinate(physeq = GV_bacteria_g_rar, method = "NMDS", 
                               distance = "bray")

P_g_b_NMDS<- 
  plot_ordination(physeq = GV_bacteria_g_rar, ordination = NMDS_g_bacteria_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Grease bacteria NMDS rarefied")
```

### PCoA Archaea
```{r}
bray_pcoa_g_archaea_rar <- ordinate(
  physeq = GV_archaea_g_rar,
  method = "PCoA", 
  distance = "bray")

P_g_a_pcoa<-
  plot_ordination(physeq = GV_archaea_g_rar, ordination = bray_pcoa_g_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Grease archaea PCoA rarefied")
```

### NMDS Archaea
```{r}
NMDS_g_archaea_rar <- ordinate(physeq = GV_archaea_g_rar, method = "NMDS", 
                                distance = "bray")

P_g_a_NMDS<-
  plot_ordination(physeq = GV_archaea_g_rar, ordination = NMDS_g_archaea_rar,
                label = "Day_ID") +
  geom_point(aes(color = as.factor(OLR_VS)), size = 3,
             stroke = 1)+
  theme_classic()+
  scale_color_manual(values = c("blue","green","red","pink","orange"))+
  theme(legend.title=element_blank(),
        legend.text = element_text(size = 10, face = "bold"))+
  labs(title = "Grease archaea NMDS rarefied")
```

### combine
```{r}
(P_g_b_NMDS | P_g_b_pcoa) / (P_g_a_NMDS | P_g_a_pcoa)
```

## Combine Bacteria PCoA plots
```{r}
knitr::opts_chunk$set(fig.width = 20, fig.height=20)
```


```{r fig.height=9, fig.width=13, message=FALSE, warning=FALSE}
(P_p_b_pcoa | P_c_b_pcoa) / (P_g_b_pcoa | P_all_b_pcoa)
ggsave("GV_pcoa_all_interval_1.jpeg", width=15, heigh = 10, units=c("in"))

```


#Permanova

## remove OLR_VS tab for whole data
```{r}
samdata_no_olr<-subset(samdata, select=-OLR_VS)
```


## Check homogeneity for substrate
```{r}
ASV_d<-vegan::vegdist(ASVtab, method="bray")

anova(vegan::betadisper(ASV_d, samdata_no_olr$substrate))
```
#showing variance homogeneity because no significant difference- good

## Permanova on Substrate
```{r}
ASV_m<-as.matrix(ASVtab)
otu.rrpp<-RRPP::lm.rrpp(ASV_d ~ substrate,
                        data= samdata_no_olr, SS.type="III",
                        print.progress=FALSE,
                        seed="random",
                        iter=999)
```

```{r}
anova(otu.rrpp, effect.type="F",
      error=c("Residuals"))
```
## significant difference based on substrate

###Pairwise
```{r}
PW<-RRPP::pairwise(otu.rrpp, groups=samdata_no_olr$substrate)
summary(PW)
```

### FDR correction
```{r}
SPW<-summary(PW)
pvals.sub<-SPW$pairwise.table$P # get pvalue matrix

# FDR correction
fdr.pvals.sub<-p.adjust(pvals.sub,"fdr")
fdr.pvals.sub<-matrix(fdr.pvals.sub, nrow=3, ncol=3, byrow= FALSE)
pval_col<-c("Carbohydrate","Grease","Protein")
pval_row<-c("Carbohydrate","Grease","Protein")
rownames(fdr.pvals.sub)<-pval_row
colnames(fdr.pvals.sub)<-pval_col

letters.sub<-multcompLetters(fdr.pvals.sub < 0.05)$Letters 
```

###dispersion check
```{r}
disp_sub<-betadisper(ASV_d, samdata$substrate, type="centroid")
anova(disp_sub)
```
### plot
```{r}
plot(disp_sub, hull = FALSE, ellipse = TRUE) # Shows group spread
```

```{r}
plot(disp_sub, axes = c(1,3)) # View PC1 vs PC3 instead of PC1 vs PC2
```



### check with vegan adonis
```{r}
samdata_no_olr.d<-data.frame(samdata_no_olr)
adonis_sub<- adonis2(ASV_d~substrate, data=samdata_no_olr.d)
```

### check with pairwise adonis
```{r}
PW_adonis_sub<-pairwise.adonis(ASV_d, samdata_no_olr.d$substrate, p.adjust.m = "fdr", perm=999)
```


# Signficant differences within substrate categories
#### looking at OLR_vs and recovery status

# Protein- OLR_vs
```{r}
#check homogeneity
ASV_p_d<-vegan::vegdist(ASVtab_p, method="bray")

anova(vegan::betadisper(ASV_p_d, samdata_p$OLR_VS))
```
### rrpp
```{r}
ASV_m_p<-as.matrix(ASVtab_p)
otu.rrpp_p_olr<-RRPP::lm.rrpp(ASV_p_d ~ OLR_VS+pH+Ammonia,
                        data= samdata_p, SS.type="III",
                        print.progress=FALSE,
                        seed="random",
                        iter=999)
```

```{r}
anova(otu.rrpp_p_olr, effect.type="F")
```
### Pairwise
```{r}
#pairwise
PW_p_olr<-RRPP::pairwise(otu.rrpp_p_olr, groups=samdata_p$OLR_VS)
summary(PW_p_olr)
```

### adonis

```{r}
set.seed(51426)
samdata_p.df<-data.frame(samdata_p)
adonis_p<-adonis2(ASV_p_d~OLR_VS, data=samdata_p.df)
```

### adonis with other parameters
```{r}
adonis_p_co<- adonis2(ASV_p_d~OLR_VS+pH+Ammonia, data=samdata_p.df, by="margin")
#OLR showing statistical significance, could also categorize continous variables but considering not much variability in pH not sure how easy or useful this would be. excluded TVFA because it seems to correlate to pH. 
```

```{r}
adonis_p_nh3<- adonis2(ASV_p_d~Ammonia, data=samdata_p.df)
```

# Carbohydrate OLR_vs and Recovery
### check dispersion on OLR
```{r}
ASV_c_d<-vegan::vegdist(ASVtab_c, method="bray")

anova(vegan::betadisper(ASV_c_d, samdata_c$OLR_VS))
```
### check dispersion of recovery
```{r}

anova(vegan::betadisper(ASV_c_d, samdata_c$Recovery))
# both dispersions are not significant so they are good
```
### adonis
```{r}
set.seed(51426)
samdata_c.df<-data.frame(samdata_c)
adonis_c_olr<-adonis2(ASV_c_d~OLR_VS, data=samdata_c.df)
#showing significance based on OLR
```

```{r}
adonis_c_rec<-adonis2(ASV_c_d~Recovery, data=samdata_c.df)
#showing signficant difference based on OLR
```

```{r}
adonis_c_co<-adonis2(ASV_c_d~OLR_VS+pH+Ammonia, data=samdata_c.df, by="margin") 
```

### pairwse adonis

```{r}
set.seed(51426)
PW_adonis_c_olr<-pairwise.adonis(ASV_c_d, samdata_c.df$OLR_VS, p.adjust.m = "fdr", perm=999)
```


### rrpp
```{r}
ASV_m_c<-as.matrix(ASVtab_c)
otu.rrpp_c_co <- lm.rrpp(ASV_c_d ~ OLR_VS*Recovery + pH + Ammonia, 
                          data = samdata_c, 
                          SS.type = "III")
```

```{r}
anova(otu.rrpp_c_co, effect.type="F")
```

# Grease OLR_vs and Recovery
##check dispersion on OLR
```{r}
ASV_g_d<-vegan::vegdist(ASVtab_g, method="bray")

anova(vegan::betadisper(ASV_g_d, samdata_g$OLR_VS))
```

## check dispersion of Recovery
```{r}
anova(vegan::betadisper(ASV_g_d, samdata_g$Recovery))
```

##adonis
```{r}
set.seed(51426)
samdata_g.df<-data.frame(samdata_g)
adonis_g_olr<-adonis2(ASV_g_d~OLR_VS, data=samdata_g.df)
```

```{r}
adonis_g_rec<-adonis2(ASV_g_d~Recovery, data=samdata_g.df)
```

```{r}
adonis_g_co<-adonis2(ASV_g_d~OLR_VS+Recovery+pH+Ammonia, data=samdata_g.df, by="margin")
```

## rrpp
```{r}
ASV_m_g<-as.matrix(ASVtab_g)
otu.rrpp_g_co <- lm.rrpp(ASV_g_d ~ OLR_VS*Recovery + pH + Ammonia, 
                          data = samdata_g, 
                          SS.type = "III")
```

```{r}
anova(otu.rrpp_g_co, effect.type="F")
```
```{r}
otu.rrpp_g_olr <- lm.rrpp(ASV_m_g ~ OLR_VS, 
                          data = samdata_g, 
                          SS.type = "III")

anova(otu.rrpp_g_olr, effect.type="F")
```


# code from gemini to treat ph ammonia and tvfa as covariates
```{r}
# Multi-factor RRPP model incorporating treatments and covariates
mod_bioreactor <- lm.rrpp(ASV_d ~ substrate + pH+ Ammonia+ OLR_num, 
                          data = samdata_no_olr, 
                          SS.type = "III") # Crucial for unique variance partitioning
```


```{r}
anova(mod_bioreactor, effect.type="F") #leave of error so it will automatically run residuals
```

#dbRDA
## Protein
```{r}
set.seed(51426)
dbrda_p<-dbrda(ASV_p_d~ OLR_num+pH, data=samdata_p.df)
vif.cca(dbrda_p)
anova(dbrda_p)

margin_p<-anova(dbrda_p, by="margin", permutations=999)
print(margin_p)
```
```{r}
plot(dbrda_p)
```



## Carbohydrate
```{r}
set.seed(51426)
dbrda_c<-dbrda(ASV_c_d~ Ammonia+pH+OLR_VS, data=samdata_c.df, method="bray")
vif.cca(dbrda_c)
anova(dbrda_c)

margin_c<-anova(dbrda_c, by="margin", permutations=999)
print(margin_c)
```
```{r}
plot(dbrda_c)
```


## Grease
```{r}
set.seed(51426)
dbrda_g<-dbrda(ASV_g_d~ Ammonia+pH+OLR_VS, data=samdata_g.df)
vif.cca(dbrda_g)
summary(dbrda_g)

margin_g<-anova(dbrda_g, by="margin", permutations=999)
print(margin_g)
```

```{r}
plot(dbrda_g)
```


# vegan cca
##protein
```{r}
cca_p<-cca(ASVtab_p~ OLR_num+pH, data=samdata_p_std)
vif.cca(cca_p)

summary(cca_p)
anova(cca_p, by="margin")
plot(cca_p, display = c("wa","bp","cn")
  
P_cca_p<-autoplot(cca_p, display=c("wa","bp"), geom= "text")+
  theme_classic()+
  theme(legend.position="none")+
  labs(title="A")
```

```{r}
plot(cca_p, type="n")
points(cca_p, display="wa", pch=21, col="blue", bg="blue", cex=2)
text(cca_p, display="bp", col="black", cex=2)

```



## carbohyrdate
```{r}
cca_c<-cca(ASVtab_c~ OLR_num+pH+Ammonia, data=samdata_c_std)
vif.cca(cca_c)

summary(cca_c)
anova(cca_c, by="margin")
P_cca_c<-plot(cca_c, display = c("wa","bp","cn"))
P_cca_c<-autoplot(cca_c, display=c("wa","bp"), geom= "text")+
  theme_classic()+
  theme(legend.position="none")+
  labs(title="B")
```
## grease
```{r}
cca_g<-cca(ASVtab_g~ OLR_num+pH+Ammonia, data=samdata_g_std)
vif.cca(cca_g)

summary(cca_g)
anova(cca_g, by="margin")

P_cca_g<-autoplot(cca_g, display=c("wa","bp"), geom= "text")+
  theme_classic()+
  theme(legend.position="none")+
  labs(title="C")
```

# combine CCA plots
```{r}

(P_cca_p|P_cca_c|P_cca_g)
ggsave("GV_cca_plots.jpeg", width=20, heigh = 15, units=c("in"))
```

```{r}
#this is a test#
```

```{r}
#this is a second test#
```

